My title

### 2024-11-26 paper/paper.qmd
draft

---
title: "My title"
subtitle: "My subtitle if needed"
author: Cristina Su Lam
thanks: "Code and data are available at: https://github.com/cristinaasu/Spotify_Analysis"
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
toc: true
fig_caption: yes
number-sections: true
format: pdf
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false
# Load required libraries
library(tidyverse)
library(ggplot2)
library(hexbin)
library(arrow)
```

```{r}
# Load dataset 
analysis_data <- read_parquet("data/02-analysis_data/analysis_data.parquet")

```

@@

Overview paragraph

@@

Telegraphing paragraph: The remainder of this paper is structured as follows. @sec-data....

@@

Overview text

@@

Some paragraphs about how we go from a phenomena in the world to an entry in the dataset.

## Outcome variable

Add graphs, tables and text. Use sub-sub-headings for each outcome variable or update the subheading to be singular.

Some of our data is of penguins (@fig-bills), from @palmerpenguins.

```{r}
#| label: fig-valence
#| fig-cap: Distribution of Valence
#| echo: false

ggplot(analysis_data, aes(x = Valence)) +
  geom_histogram(binwidth = 0.05, fill = "blue", color = "black", alpha = 0.7) +
  labs(title = "Distribution of Valence", x = "Valence", y = "Count") +
  theme_minimal()
```

Talk more about it.

And also planes (@fig-planes). (You can change the height and width, but don't worry about doing that until you have finished every other aspect of the paper - Quarto will try to make it look nice and the defaults usually work well once you have enough text.)

```{r}
#| label: fig-planes
#| fig-cap: Relationship between wing length and width
#| echo: false
#| warning: false
#| message: false

analysis_data <- read_csv(here::here("data/02-analysis_data/analysis_data.csv"))

analysis_data |> 
  ggplot(aes(x = width, y = length)) +
  geom_point(alpha = 0.8) +
  theme_minimal() +
  labs(x = "Wing width (mm)",
       y = "Wing length (mm)")
```

Talk way more about it.

## Predictor variables

Add graphs, tables and text.

Use sub-sub-headings for each outcome variable and feel free to combine a few into one if they go together naturally.

>>>>

---
title: "My title"
subtitle: "My subtitle if needed"
author: Cristina Su Lam
thanks: "Code and data are available at: https://github.com/cristinaasu/Spotify_Analysis"
date: today
date-format: long
abstract: "This analysis explores the relationship between various musical and contextual features with the emotional tone (Valence) of songs on Spotify. Using a dataset of songs with features such as tempo, danceability, and acousticness, alongside contextual data like mean temperature, we aim to predict Valence—a measure of positivity in the music. The study applies multiple linear regression models, evaluates performance metrics, and identifies influential predictors. Results show that predictors like danceability and energy strongly relate to Valence, while contextual variables like temperature also provide moderate explanatory power. These findings highlight the potential for combining musical and environmental data to analyze listener sentiment and mood-related trends in music[need to be more detail about findings]."
toc: true
fig_caption: yes
number-sections: true
format: pdf
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false
library(tidyverse)
library(ggplot2)
library(hexbin)
library(arrow)
library(kableExtra)
library(knitr)
```

```{r}
# Load dataset 
analysis_data <- read_parquet("../data/02-analysis_data/analysis_data.parquet")
mlm_model <- readRDS("../models/mlm_model.rds")
```

@@

Valence, a measure of emotional tone in music, is pivotal for understanding how listeners perceive and engage with songs. Ranging from 0 (negative or sad) to 1 (positive or happy), Valence provides insights into the emotive qualities of a track. This study aims to model and predict Valence using a blend of musical features such as tempo, danceability, and acousticness, alongside contextual variables like mean temperature.

Music serves as a universal language, evoking emotions and creating connections. By analyzing how intrinsic musical properties interact with external factors, such as environmental conditions, we can uncover patterns that shape listener sentiment. This approach not only deepens our understanding of the factors influencing emotional tone but also offers practical applications for playlist curation and music recommendation systems.

The estimand of interest is Valence, as it encapsulates the core emotional response elicited by music. Through multiple linear regression (MLR) models, we identify the most significant predictors of Valence and assess their explanatory power. This approach highlights the interplay between quantitative attributes of music and listener experiences, aiming to improve the accuracy of mood prediction models.

In this analysis, we seek to bridge the gap between musical analytics and contextual data, contributing to the broader field of musicology and data-driven sentiment analysis. By refining predictive methods, we aim to better understand the emotional landscape of modern music and its relationship with the listener's environment.

[Overview paragraph

@@

Telegraphing paragraph: The remainder of this paper is structured as follows. @sec-data....]

@@

The dataset contains key variables spanning musical features, environmental data, and outcomes. Below is a summary of numerical variables:
```{r}
#| label: tbl-summary
#| tbl-cap: Summary Statistics
#| echo: false
#| warning: false
#| fig-align: center
summary_table <- analysis_data %>%
  reframe(
    Variable = c("Valence", "Mean Temp", "Danceability", "Energy", "Acousticness", "Tempo"),
    Mean = c(
      mean(Valence, na.rm = TRUE),
      mean(`Mean Temp`, na.rm = TRUE),
      mean(Danceability, na.rm = TRUE),
      mean(Energy, na.rm = TRUE),
      mean(Acousticness, na.rm = TRUE),
      mean(Tempo, na.rm = TRUE)
    ),
    SD = c(
      sd(Valence, na.rm = TRUE),
      sd(`Mean Temp`, na.rm = TRUE),
      sd(Danceability, na.rm = TRUE),
      sd(Energy, na.rm = TRUE),
      sd(Acousticness, na.rm = TRUE),
      sd(Tempo, na.rm = TRUE)
    ),
    Min = c(
      min(Valence, na.rm = TRUE),
      min(`Mean Temp`, na.rm = TRUE),
      min(Danceability, na.rm = TRUE),
      min(Energy, na.rm = TRUE),
      min(Acousticness, na.rm = TRUE),
      min(Tempo, na.rm = TRUE)
    ),
    Max = c(
      max(Valence, na.rm = TRUE),
      max(`Mean Temp`, na.rm = TRUE),
      max(Danceability, na.rm = TRUE),
      max(Energy, na.rm = TRUE),
      max(Acousticness, na.rm = TRUE),
      max(Tempo, na.rm = TRUE)
    )
  ) %>%
  mutate(
    Mean = round(Mean, 3),
    SD = round(SD, 3),
    Min = round(Min, 3),
    Max = round(Max, 3)
  )
kable(summary_table, format = "pipe")
```
These metrics reveal diverse song features and moderate variability in environmental temperature, offering rich explanatory potential for Valence.

@@

The data used in this analysis were collected from Spotify’s public API, which provides detailed metadata for songs. Each track is associated with multiple audio features and contextual information. The data was preprocessed to ensure consistency, with missing or erroneous values removed prior to analysis.

Valence (Outcome): Extracted directly from Spotify’s API, Valence is measured on a scale from 0 to 1, indicating the positivity or happiness conveyed by a song. This variable serves as the primary outcome of interest.

Musical Features:

Danceability: Quantifies how suitable a track is for dancing, based on a combination of tempo, rhythm stability, and beat strength. Values range from 0 to 1.

Energy: Measures the intensity and activity level of a track, combining attributes like loudness and dynamic range. Higher values indicate more energetic tracks.

Acousticness: Represents the likelihood of a track being acoustic, with values closer to 1 reflecting higher acoustic qualities.

Tempo: Captured in beats per minute (BPM), tempo describes the overall speed of a track. This feature was extracted directly from Spotify’s analysis tools.

Contextual Feature:

Mean Temp: Weather data was integrated using external sources and mapped to listening periods. This feature reflects the average daily temperature (°C) corresponding to the time a track was popular or listened to.

The integration of audio and contextual data allowed for a comprehensive analysis of factors influencing Valence, bridging the gap between intrinsic musical properties and external environmental conditions.

[Some paragraphs about how we go from a phenomena in the world to an entry in the dataset.]

## Outcome variable
The distribution of Valence shows a relatively uniform spread, with some concentration near the mid-range. This reflects the diversity of emotional tones in the dataset. The histogram of Valence illustrates this distribution, where most tracks lie between moderate to high levels of positivity.
```{r}
#| label: fig-valence
#| fig-cap: Distribution of Valence
#| echo: false
#| warning: false
#| fig-align: center

ggplot(analysis_data, aes(x = Valence)) +
  geom_histogram(binwidth = 0.05, fill = "blue", color = "black", alpha = 0.7) +
  labs(x = "Valence", y = "Count") +
  theme_minimal()
```

## Predictor variables
### Mean Temperature

The average temperature over time demonstrates a seasonal decline, ranging from approximately 25°C in the summer to below 10°C in late autumn. This variable adds a contextual layer, capturing external influences that may correlate with song Valence and listening behavior.
```{r}
#| label: fig-temp
#| fig-cap: Average temperature Over Time
#| echo: false
#| warning: false
#| fig-align: center
ggplot(analysis_data, aes(x = Date, y = `Mean Temp`)) +
  geom_line(color = "red") +
  labs(title = "Average Temperature Over Time", x = "Date", y = "Temperature (°C)") +
  theme_minimal()
```
### Tempo
The distribution of Tempo reveals a central peak around 120 BPM, with a wider spread for slower and faster songs. This suggests a balanced mix of tempos, with a slight preference for moderate-speed tracks. Tempo’s role in influencing Valence is analyzed alongside other musical features.

```{r}
#| label: fig-tempo
#| fig-cap: Distribution of Tempo
#| echo: false
#| warning: false
#| fig-align: center
ggplot(analysis_data, aes(x = Tempo)) +
  geom_histogram(binwidth = 10, fill = "darkgreen", color = "black") +
  labs(title = "Distribution of Tempo", x = "Tempo (BPM)", y = "Frequency") +
  theme_minimal()
```

### Danceability, Energy, and Acousticness

The comparative histograms of Danceability, Energy, and Acousticness highlight their distinct distributions:

Danceability: Peaks near 0.6 to 0.8, suggesting that most songs are moderately to highly danceable.

Energy: Similarly peaks in the mid-range, reflecting a tendency towards active, engaging tracks.

Acousticness: Displays a skewed distribution, with most tracks having low acoustic properties, indicating a focus on electronic or produced music styles.

Each of these variables provides unique explanatory power, contributing to the overall understanding of what drives the emotional tone in music.

```{r}
#| label: fig-tempo
#| fig-cap: Distribution of Tempo
#| echo: false
#| warning: false
#| fig-align: center
analysis_data %>%
  select(Danceability, Energy, Acousticness) %>%
  pivot_longer(cols = everything(), names_to = "Variable", values_to = "Value") %>%
  ggplot(aes(x = Value, fill = Variable)) +
  geom_histogram(binwidth = 0.05, alpha = 0.6, position = "identity", color = "black") +
  facet_wrap(~Variable, scales = "free") +
  labs(
    title = "Distribution of Danceability, Energy, and Acousticness",
    x = "Value",
    y = "Frequency"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```
